---
title: "Manipulating data.frame"
date: "29 August 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

So far we've learned some basics of manipulating data.frame objects and how 
we can use conditional operations to subset vectors. Now, we're going to combine 
these skills to make more advanced table manipulations. We will learn:

* How to filter a data.frame based on certain conditions
* How to summarise data across groups

This session is adapted from 
[this lesson](http://www.datacarpentry.org/R-ecology-lesson/03-dplyr.html) 
of the Data Carpentry materials. 

We will continue using the `surveys` dataset and the `tidyverse` package:

```{r, message=FALSE}
# Load the package
library(tidyverse)

# Read the data
surveys <- read_csv("data/surveys_data_short.csv")
```

# Advanced manipulations with dplyr

`dplyr` is a package that comes included in the `tidyverse` package that we've 
been using. It contains many functions to manipulate data.frame objects.

We can think of dplyr functions as verbs. We will talk about the following:

* `select` - to select specific columns
* `filter` - to filter rows
* `mutate` - to modify existing columns or create new ones
* `group_by` - to define groups within our data.frame
* `summarise` - to apply functions to groups
* `count` - to count observations in groups

Every `dplyr` function sticks to these rules:

* **The first argument (input) of every `dplyr` function is a data.frame.**
* **The output of every `dplyr` function is a data.frame**

## Subset a data.frame with `filter()`

Earlier, we've written a condition to obtain the weights of individuals of the 
"PH" species that have a hindfoot length greater than 25mm:

```{r}
# Make a logical operator for which the species ID is equal to "PH"
ph_hindfoot25 <- surveys$species_id == "PH" & surveys$hindfoot_length > 25

# Subset the weights based on that condition
surveys$weight[ph_hindfoot25]
```

But what if we wanted not just a vector of those weights, but a subset of the 
table that fullfilled that condition?

With `filter()` we could do it like this:

```{r}
filter(surveys, species_id == "PH" & hindfoot_length > 25)
```

Notice two things from here:

* The first argument of the function is the data.frame, as we said before
* Then we introduce the condition, but we don't need to use `$` anymore, because 
the function knows that those are columns of the data.frame we gave to it.

**This applies to every `dplyr` function.**


## Create/modify data.frame variables with `mutate()`

A very common manipulation to make in data.frame objects is to create new 
variables or modify existing ones. We can do this with the `mutate()` function.

For example, if we wanted to create a new variable that contained the weight in 
Kg:

```{r}
mutate(surveys, weight_kg = weight / 1000)
```

We can also modify/create more than one variable, separating them by `,`:

```{r}
mutate(surveys, weight_kg = weight / 1000, hindfoot_cm = hindfoot_length / 10)
```

If we use the name of an existing variable, it will modify it instead. For example:

```{r}
mutate(surveys, weight = weight / 1000, hindfoot_cm = hindfoot_length / 10)
```

> Add a new column to the surveys data.frame called `heavy` containing logical 
values of whether or not the individual's weight is above or equal to 200g

```{r}
# put your code here!
```


## Chain commands with `%>%`

Now let's say that we wanted to do several manipulations in our data.frame:

* Create a new column containing weight in Kg
* Filter our data so that it only contains individuals collected in the 90's
* Retain only two columns from our table: `species_id` and `year`

Based on what we've learned so far, we could for example do the following:

```{r}
# Create the new variable with mutate
filtered_surveys <- mutate(surveys, weight_kg = weight/1000)

# Filter this new table to retain only collections in the 90's
filtered_surveys <- filter(filtered_surveys, year >= 1990 & year < 2000)

# Use select to retain the two columns of interest
filtered_surveys <- select(filtered_surveys, year, species_id)
```

Someone might even attempt to write this code like this (can you read it!?!):

```{r}
# Mutate, filter and select all nested within each other
filtered_surveys <- select(filter(mutate(surveys, weight_kg = weight/1000), year >= 1990 & year < 2000), year, species_id)
```

The above two solutions have some disadvantages. In the first case we had to create 
several intermediate variables before we got to the final result that we were 
interested in. In the second case, it's clearly very difficult code to read.

`dplyr` allows to "chain" several commands using a special "pipe" function `%>%`.
This is how the above code could have been written using this trick:

```{r, eval = FALSE}
surveys %>% 
  mutate(weight_kg = weight/1000) %>% 
  filter(year >= 1990 & year < 2000) %>% 
  select(year, species_id)
```

Essentially, what the `%>%` "pipe" does is send the output of the expression on its 
left and passes it as the **first** input of the function on its right. And this 
works very well with `dplyr` functions, because of the two rules that every one 
of its functions follows: first argument is a table; output is a table.

>Create a new data frame from the survey data that meets the following criteria: 
>
>* contains only the species_id column and a new column called hindfoot_half
>* hindfoot_half will contain values that are half the hindfoot_length values
>* In this hindfoot_half column, there are no NAs and all values are less than 30

```{r}
# put your code here!
```


## Summarise data with `group_by()` and `summarise()`

One of our purposes from the beggining of this R session was to summarise our 
data into a smaller table with the mean weight and hindfoot length of each species 
and each sex. 

This kind of operation can be achieved by combining two `dplyr` functions 
together: `group_by()` and `summarise`. Here is how it works:

```{r}
surveys %>% 
  group_by(genus) %>% 
  summarise(mean_weight = mean(weight))
```

The function `group_by` does not change our initial table, but it adds to it some 
information that we want to do some operations within groups. The `summarise` 
function then works within each group, by applying the functions we ask it to. 

Like with `mutate()`, with `summarise()` we can also create more than one variable 
at once, for example:

```{r}
surveys %>% 
  group_by(genus) %>% 
  summarise(mean_weight = mean(weight), max_weight = max(weight))
```

We can also group a table by more than one group, for example:

```{r}
surveys %>% 
  group_by(genus, species) %>% 
  summarise(mean_weight = mean(weight), max_weight = max(weight))
```

> Using `group_by()` and `summarize()` create a new table called `surveys_summary` 
with the mean, minimum and max hindfoot length and weight for each `species_id` and `sex`
>
> Create a scatter plot showing the relationship between the average weight and 
hindfoot length of each species

```{r}
# put your code here!
```


## Count occurrences in groups

Often, it's useful to count how many occurrences of each variable (or variables) 
we have in a dataset. This can be achieved with the `count()` function, which 
works like this:

```{r}
# Count occurrences of each genus
surveys %>% count(genus)

# Count occurrences of each genus and species
surveys %>% count(genus, species)
```


# Saving data to a file

After all our data manipulations, we might want to save some new table into a file.

Let's say that we were to remove all the missing data from our data to share it 
with a collaborator (or later use in R ourselves):

```{r}
# Remove missing values from the data
surveys_clean <- surveys %>% 
  filter(!is.na(hindfoot_length))
```

In the same way that there is a `read_csv()` function, there is also a 
`write_csv()`. This is how it works:

```{r, eval = FALSE}
write_csv(surveys_clean, file = "data/surveys_clean.csv")
```

> Try opening this file in a spreadsheet!


# Some key points to remember

* `dplyr` has several useful functions to manipulate data.frame objects
* `dplyr` functions can be "chained" together using the `%>%` pipe
* We can save data.frame to text-based files using the `write_*` family of functions
