---
title: "Variables and conditionals"
date: "29 August 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this lesson we're going to learn more about different types of variables, which 
is very important in order to manipulate and query our tables (data.frame objects). 

We're going to learn:

* What types of variables there are in R (numeric, character, logical, factor)
* What are missing values and how to deal with them
* How to subset data based on conditions
* How to apply this knowledge to subset data.frame objects

We will continue using the `surveys` dataset and the `tidyverse` package:

```{r, message=FALSE}
# Load the package
library(tidyverse)

# Read the data
surveys <- read_csv("data/surveys_data_short.csv")
```


## Variable types

As you might already suspect, our `surveys` data is composed of variables that 
are of a different nature. Some are descriptive (like `species_id` or `genus`) 
whereas others are of a numerical nature (such as `weight` and `height`).

A very useful function to get a quick summary of our data is called `summary()`:

```{r}
summary(surveys)
```

> In the output above, the summary looks different for the different variables. 
Does this make sense considering what type of variable these might be?


### Numeric and character variables

We can learn more about the contents of _data.frame_ objects with the `glimpse()` 
function:

```{r}
glimpse(surveys)
```

From the output above we see that our variables are of different "types", some are 
marked as `int` and others as `chr`. These stand, respectively, for "integer" and 
"character" and refer to the fact that these two are different types of data. 

Let's see some examples of different types of vector:

```{r}
# Numeric vector (integer)
x <- 1:4

# Numeric vector (with decimal points)
y <- c(0.1, 0.2, 3)

# Character vector
z <- c("cat", "dog", "horse")

class(x)
class(y)
class(z)
```

For all practical purposes, "integer" and "numeric" vectors are identical. They 
are both numeric in nature and can be used in arithmetic operations. Characters, 
however cannot:

```{r, error = TRUE}
mean(x)
mean(y)
mean(z) # does not work
x + y
x + z # does not work
```


## Missing values

In the `summary()` output we obtained above, we can see there is a count of `NA` 
values for `hindfoot_length`. 

```{r}
# First 10 values of hindfoot_length
surveys$hindfoot_length[1:20]
```

These represent "missing data", which R handles in a special way. 

Let's see some examples:

```{r}
# Create a vector of numbers including some NA values
x <- c(2, 3, 1, 1, NA, 10, 20)

mean(x)
sd(x)
```

If we run the functions `mean()` and `sd()` on that vector, we get back `NA`! 
This is because the default behaviour of those functions is to not proceed with 
their computation if there are any missing values in the data. If we do want them 
to calculate the mean, excluding the missing values, we have to say so explicitly:

```{r}
mean(x, na.rm = TRUE)
sd(x, na.rm = TRUE)
```

We could know about this option by looking at the help `?mean` and `?sd`.

> Calculate the mean and standard deviation of weights and hindfoot lenghts

```{r}
# put your code here!
```


## Conditionals

Let's say we wanted to know how many of our sampled individuals weight more than 
200g. To answer this question we need to learn about how to query our data based 
on some condition (in this case that the values of weight are greater than 100). 

To this end we need to learn about another type of variable: logical.

### Logical variables

There is another important variable type called a "logical". This is a type of 
variable that stores the values `TRUE` and `FALSE`. 

These variables are extremely useful, because we can use them to subset other 
vectors. See this example:

```{r}
# Note they are all uppercase with no quotes
logical_vector <- c(TRUE, FALSE, FALSE, TRUE)

x
x[logical_vector]
```

Using a logical vector within `[]` returns only the values that are `TRUE`. In 
the above example this was not very useful, because we specified the logical values 
"by hand", but consider this other case:

```{r}
# Create a vector
x <- c(10, 0.2, 20, -10, 5)
# Create a logical vector where values in x are above 2
above2 <- x > 2

# Original vector
x
# Condition
above2
# A subset the original vector
x[above2]
```

In this case, the logical vector was constructed using a condition stipulated by 
us and based on the data itself. 

We could easily find out how many of those cases fulfill our condition by using the 
`sum()` function, which for logical vectors will sum the cases containing the 
value `TRUE` (you can think of it as `TRUE = 1` and `FALSE = 0`):

```{r}
sum(above2)
```

> Based on this, we can now answer our question of how many individuals weight 
more than 200g? Hint: first create a logical vector called `weight200` that will 
contain values "TRUE" and "FALSE". Then use that vector to subset the `weight` 
column of the table using `[]`.

```{r}
# Put your code here!
```


### Conditional operators

In the example above, we managed to subset a vector based on a condition stipulated 
by us. There are many "operators" to perform this, listed below.

Relational operators:

* `==` is equal to
* `!=` is different from
* `%in%` is contained in
* `>` is greater than
* `>=` is greater than or equal to

Logical operators:

* `&` AND
* `|` OR

Combined, these can be used to perform very complex and customised operations. 

For example, here is how we could obtain the weights for the "PH" species in our 
`surveys` table that have a hindfoot length greater than 25:

```{r}
# Make a logical operator for which the species ID is equal to "PH"
ph_hindfoot25 <- surveys$species_id == "PH" & surveys$hindfoot_length > 25

# Subset the weights based on that condition
surveys$weight[ph_hindfoot25]
```

Or a condition that evaluates cases where the hindfoot length is either above 
40mm or below or equal to 10mm:

```{r, eval = FALSE}
surveys$hindfoot_length <= 10 | surveys$hindfoot_length > 40
```


### Dealing with NA values

There are some functions that also return logical vectors. One of the most 
useful ones is called `is.na()`, which evaluates if the values in a vector are 
missing or not:

```{r}
x <- c(1, 2, NA, 5, 10)

# Return true if the value is missing
is.na(x)
```

We can "negate" a logical vector by using `!`:

```{r}
# Return true if the value is NOT missing
!is.na(x)
```

This can be very handy. For example, if we wanted individuals with hindfoot 
length above 35mm, we obtain many `NA` values:

```{r}
surveys$hindfoot_length[surveys$hindfoot_length > 35]
```

We can use the `is.na()` function to solve this problem, like so:

```{r}
surveys$hindfoot_length[surveys$hindfoot_length > 35 & !is.na(surveys$hindfoot_length)]
```

We combine the two conditions into one:

* hindfoot lengths above (`>`) 35mm

AND

* values of hindfoot length that are NOT missing (`!is.na`)

> How many values of weight are there for the species "DM" with hindfoot length 
above 40mm?
> How many values of hindfoot length are either above 40mm or below 10mm?

```{r}
# put your code here!
```


## Checking data variables

Knowing what kind of variables our data is composed of and whether or not 
there are any missing values is a **very important** step in data analysis!

There are two very convenient functions called `vis_dat()` and `vis_miss()` in 
the package `visdat` that can be used to get a graphical representation of 
data in a table. They are particularly useful for big datasets like this one:

```{r}
library(visdat)
vis_dat(surveys)
vis_miss(surveys)
```

Learn more about these functions from the 
[package's website](https://cran.r-project.org/web/packages/visdat/vignettes/using_visdat.html)


## factor variables

So far, we've seen **numeric**, **character** and **logical** vectors. 

There is another type of vector, which is used to represent categorical data and 
is called **factor**.

We will learn later in the course how factors are important for statistical 
analysis. For now, consider these two vectors:

```{r}
char_vector <- c("low", "medium", "high", "high", "low", "low")
fctr_vector <- factor(char_vector)
```

> How are they different?

```{r}
char_vector
fctr_vector
```

While both contain the same values, factors have a new attribute called **levels** - 
the unique values that occur in the vector:

```{r}
levels(fctr_vector) # the levels in the factor
nlevels(fctr_vector) # the number of levels
```

Many R functions treat factors differently from character vectors. For example,
the `summary()` function that we've already used:

```{r}
summary(char_vector)
summary(fctr_vector)
```

You cannot add values that are not listed in the levels of a factor:

```{r}
char_vector[1] <- "very low"
char_vector

fctr_vector[1] <- "very low"
fctr_vector
```

And you can have levels with no observations:

```{r}
fctr_vector <- fctr_vector[fctr_vector != "medium"]
summary(fctr_vector)
```

The order of these levels is alphabetical by default, but it can be changed:

```{r}
fctr_vector <- factor(char_vector, levels = c("low", "medium", "high"))
levels(fctr_vector)
```

"Behind the scenes" R encodes each level of a factor as an integer number.

```{r}
char_vector
as.numeric(fctr_vector)
```

So be careful when converting factors that look like numbers! 

```{r}
x <- factor(c(2, 4, 5, 2, 2))
x
as.numeric(x)
```

One way to solve this problem is doing the following:

```{r}
# First convert to a character type and then to numeric type
as.numeric(as.character(x))
```

See more about factors [here](http://www.datacarpentry.org/R-ecology-lesson/02-starting-with-data.html#factors).


# Some key points to remember 

* There are different types of variables in R: numeric, character, logical, 
factor.
* Always check what type of variables you have in a data.frame.
    + for example with `vis_dat()` from the `visdat` package
* Always check for missing values in your data.
    + for example with `vis_miss()` from the `visdat` package
* You can subset data based on certain conditions using "conditional operators" 
(`>`, `<`, `>=`, `<=`, `==`, `!=`, `&`, `|`).
* Factors are special kinds of categorical variables important for statistical 
analysis (more on Days 2 and 3 of the course).




