---
title: Descriptive analyses and basic plotting in R
author: Aaron Lun, Catalina Vallejos
date: 4 April 2017
output: 
  html_document:
    fig_caption: false
---

```{r, echo=FALSE, results="hide"}
# Make compilation stop upon error:
library(knitr)
opts_chunk$set(error=FALSE)
```

# Introducing (R)markdown

Use triple backticks to indicate code environments:

```{r}
a <- 1
print(a)
```

Set `eval=FALSE` to show the code, but not run it:

```{r, eval=FALSE}
a <- 2
print(a)
```

Set `echo=FALSE` to run the code, but not show it:

```{r, echo=FALSE}
a <- 3
print(a)
```

Set `results="hide"` to hide the results:

```{r, results="hide"}
a <- 4
print(a)
```

For plots, set `fig.width` and `fig.height` to change the dimensions of the output plot:

```{r, fig.width=10, fig.height=6}
hist(rnorm(10000), xlab="X", main="Normal distribution")
```

See `?opts_chunk` and http://yihui.name/knitr/options#chunk_options for more details.

# Exploring the `cars` dataset

## Initial examination of the data

Let's look at one of the datasets that come with the R installation.
From the documentation in `?cars`:

> The data give the speed of cars and the distances taken to stop.
> Note that the data were recorded in the 1920s.  

We can inspect the first few elements:

```{r}
head(cars)
```

Or the last few elements - how would we do that?

```{r}
# Put some code here!
tail(<something>) ## QUESTION ##
tail(cars) ## ANSWER ##
```

We can have a look at some summary statistics:

```{r}
summary(cars)
```
    
What functions would we use to directly calculate:

- the mean?
- the median?
- the standard deviation?
- the mininimum or maximum?
- the first or third quartiles?

```{r}
# Put some code here!
mean(cars$speed) # mean ## ANSWER ##
median(cars$speed) # median ## ANSWER ##
sd(cars$speed) # standard deviation ## ANSWER ##
min(cars$speed) # minimum ## ANSWER ##
max(cars$speed) # maximum ## ANSWER ##
quantile(cars$speed, p=0.25) # first quartile ## ANSWER ##
quantile(cars$speed, p=0.75) # third quartile ## ANSWER ##
```

## Generating some summary plots

We can make histograms of the two sets of values:

```{r}
hist(cars$speed, xlab="Speed (mph)", main="Speed distribution")
hist(cars$dist, xlab="Distance (ft)", main="Distance distribution")
```

Check out `?hist` to tune parameters, e.g., number of bars via `breaks`, fill colour with `col`.

```{r}
# Put some code here!
hist(cars$dist, breaks=<something>, col=<something>) ## QUESTION ##
hist(cars$dist, breaks=20, col="grey") ## ANSWER ##
```

(__Note:__ how do you find out what arguments can be supplied to a function?)

We can also make a scatter plot of speeds vs distances, below.
Why is the speed on the x-axis?

```{r}
#       x-coord    y-coord
plot(cars$speed, cars$dist, xlab="Speed (mph)", ylab="Distance (ft)")
```

Alternative method, using the formula notation `~`, i.e., plotting distance as a function of speed.

```{r}
#            y ~ x
plot(cars$dist ~ cars$speed)
```

Check out `?plot` and `?par` for tunable parameters.

```{r}
# Put some code here!
plot(cars$speed, cars$dist, pch=<something>, cex=<something>, col=<something>) ## QUESTION ##
plot(cars$speed, cars$dist, pch=15, cex=2, col="red") ## ANSWER ##
```

# Exploring the `chickwts` dataset

## Initial examination of the data

From the documentation in `?chickwts`:

> An experiment was conducted to measure and compare the effectiveness of various feed supplements on the growth rate of chickens.

Examine the data:

```{r}
# Put some code here!
head(<something>) ## QUESTION ##
head(chickwts) ## ANSWER ##
```

Looking at some summaries of the data:

```{r}
# Put some code here!
summary(<something>) ## QUESTION ##
summary(chickwts) ## ANSWER ## 
```

Getting the number of samples for each feed type:

```{r}
table(chickwts$feed)
```

Previous `summary` computes values across all feed types - what's missing here?

## Computing summaries for each feed type

Using `dplyr`, we can calculate summary statistics for each feed group:

```{r}
library(dplyr)
by.feed.type <- chickwts %>%
  group_by(feed) %>%
  summarise(min = min(weight),
            q1 = quantile(weight, p = 0.25),
            median = quantile(weight, p = 0.5),
            mean = mean(weight),
            q3 = quantile(weight, p = 0.75),
            max = max(weight),
            sd = sd(weight),
            n = n())

by.feed.type
```

Notice the use of the special function `n()`, which counts how many observations 
there is in each group.


## Making various types of plots

### Making a boxplot

Let's make a boxplot:

```{r}
boxplot(chickwts$weight ~ chickwts$feed, ylab="Chick weight (g)")
```

Again, using the formula notation to plot weight against the feed type.

Check out `?boxplot` for tunable parameters:

```{r}
feed.cols <- c("blue", "red", "green", "yellow", "purple", "pink") 
# Put some code here!
boxplot(chickwts$weight ~ chickwts$feed, ylab="Chickweight (g)", col=<something>) ## QUESTION ##
boxplot(chickwts$weight ~ chickwts$feed, ylab="Chickweight (g)", col=feed.cols) ## ANSWER ##
```

What do the points in the plot represent (hint: have a look at the `range` argument in `?boxplot`)?

### Making a barplot

We can also make a barplot of the mean weights for all feed types.

```{r}
barplot(by.feed.type$mean, ylab = "Mean weight (g)")
```

This would be more useful if it had labels for each bar. 
How can we do this (hint: check out `?barplot` for extra parameters)?

```{r}
# Put some code here!
barplot(by.feed.type$mean, names.arg = <something>, ylab="Mean weight (g)") ## QUESTION ##
barplot(by.feed.type$mean, names.arg = by.feed.type$feed, ylab="Mean weight (g)") ## ANSWER ##
barplot(by.feed.type$mean, names.arg = <something>, ylab="Mean weight (g)", col=<something>) ## QUESTION ## 
barplot(by.feed.type$mean, names.arg = by.feed.type$feed, ylab="Mean weight (g)", col=feed.cols) ## ANSWER ##
```

__Advanced__: how do we add error bars?
Let's compute the standard error:

```{r}
sd.wts <- by.feed.type$sd
n.obs <- by.feed.type$n
se.wts <- sd.wts/sqrt(n.obs)
```

What's the difference between `sd.wts` and `se.wts`, and which one should we use?

Can you interpret what's happening in the code below?

```{r}
mean.wts <- by.feed.type$mean
x.pos <- barplot(mean.wts, ylab="Mean weight (g)", names.arg = by.feed.type$feed, ylim=c(0, max(mean.wts)*1.1))
mean.plus.se <- mean.wts + se.wts
segments(x.pos, mean.wts, x.pos, mean.plus.se)
segments(x.pos+0.1, mean.plus.se, x.pos-0.1, mean.plus.se)
```

* The first `segments` draws the vertical part of the error bar. ## ANSWER ##
* The second `segments` draws the horizontal part of the error bar. ## ANSWER ##
* The `ylim` argument ensures that the entirety of the error bar is visible. ## ANSWER ##

# Plotting lines with the `Orange` dataset

## Initial examination of the data

This dataset examines the growth of a number of orange trees over time.

```{r}
# Put some code here!
head(<something>) ## QUESTION ##
head(Orange) ## ANSWER ##
```

Having a look at some data summaries:

```{r}
# Put some code here!
summary(<something>) ## QUESTION ##
summary(Orange) ## ANSWER ##
```

How many trees do we have?
How many data points do we have per tree?

```{r}
# Put some code here!
table(Orange$<something>) ## QUESTION ##
table(Orange$Tree) ## ANSWER ##
```

To calculate the _mean_, _standard deviation_ and _median_ for both age and circumference of each tree:

```{r}
by.tree <- Orange %>%
  group_by(Tree) %>%
  summarise(mean_age = mean(age),
            mean_circumference = <do it yourself>, ## QUESTION ##
            mean_circumference = mean(circumference), ## ANSWER ##
            sd_age = sd(age),
            sd_circumference = <do it yourself>, ## QUESTION ##
            sd_circumference = sd(circumference), ## ANSWER ##
            median_age = median(age),
            median_circumference = <do it yourself>) ## QUESTION ##
            median_circumference = median(circumference)) ## ANSWER ##
            
by.tree
```

## Making line plots

Making plots of the data:

```{r}
# Put some code here!
plot(<what is x>, <what is y>, xlab="Age (days)", ylab="Circumference (mm)") ## QUESTION ##
plot(Orange$age, Orange$circumference, xlab="Age (days)", ylab="Circumference (mm)") ## ANSWER ##
```

Each data point is linked to one at an earlier time, _for the same tree_ - how do we represent this (hint: `lines`)?

```{r}
plot(Orange$age, Orange$circumference, xlab="Age (days)", ylab="Circumference (mm)")
# Put some code here!
is.tree <- <pick all observations for tree 1> ## QUESTION ##
is.tree <- Orange$Tree==1 ## ANSWER ##
lines(Orange$age[is.tree], Orange$circumference[is.tree]) 
```

There is a way to automate this kind of repetitive task using a programming technique called a `for` loop.
Try and see if you understand what the following code does:

```{r}
plot(Orange$age, Orange$circumference, xlab="Age (days)", ylab="Circumference (mm)")
for (tree in unique(Orange$Tree)) { 
    is.tree <- Orange$Tree==tree 
    lines(Orange$age[is.tree], Orange$circumference[is.tree]) 
} 
```

__Advanced:__ Flexing R's graphical muscle:

```{r}
tree.numbers <- as.character(Orange$Tree)
unique.trees <- unique(tree.numbers)
all.colors <- rainbow(length(unique.trees))
names(all.colors) <- unique.trees 
plot(Orange$age, Orange$circumference, xlab="Age (days)", ylab="Circumference (mm)", 
    col=all.colors[tree.numbers], pch=16)
for (tree in unique.trees) {
    is.tree <- Orange$Tree==tree 
    lines(Orange$age[is.tree], Orange$circumference[is.tree], col=all.colors[tree], lwd=2)
} 
legend("topleft", col=all.colors, pch=16, lwd=2, legend=paste("Tree", unique.trees))
```

__WARNING!__ Be careful of the ordering of x-coordinates when using `lines`!
How do we fix the following code?

```{r}
x <- runif(50)
y <- (x-0.5)^2
plot(x, y)
lines(x, y)

# Need to order observations by increasing 'x'
plot(x, y) 
# Put some code here!
o <- order(<something>) ## QUESTION ## 
o <- order(x) ## ANSWER ##
print(x)
print(x[o])
lines(x[o], y[o]) 
```

# Further on the power of R graphics

A lot of graphical fine-tuning can be performed using the `par` command.
Looking at `?par` reveals a _lot_ of widely applicable graphical parameters that are not listed in the documentation of individual plotting functions.

* `mfrow`, to make multi-panel plots (advanced users may prefer `layout`)
* `mar`, to modify the margins of the plot
* `cex.axis` and `cex.label`, to change the font size of axis labels

If you get lost or confused, remember: Dr Google is your friend (if you know what to ask).

An alternative to Rstudio's export is to save R plots directly to file with various graphics devices.
For example, to save to PDF:

```{r}
pdf("my_plot.pdf")
plot(cars$speed, cars$dist)
dev.off()
```

PDFs are preferable for viewing plots in high resolution, but you can also save to PNG (e.g., if there are too many data points to store in a PDF).
Some tinkering is required to save in a decent resolution, though:

```{r, fig.keep='none'}
png("my_plot.png", width=2000, height=2000, res=300, pointsize=15)
plot(cars$speed, cars$dist)
dev.off()
```

# Session information

It's always wise to store your session information, so you know what version of R (and its packages) you ran it on.

```{r}
# Put some code here!
sessionInfo() ## ANSWER ##
```

